Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-12 manuscript.tex
Andy edits

\noindent A robust understanding of the sequence-dependent thermodynamics of DNA hybridization has enabled rapid advances in DNA nanotechnology. A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped. In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy. For a repetitive ATATATATAT sequence \noteb{not sure if we want to use 5$^\prime$-ATATATATAT-3$^\prime$ notation here as well?}, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations. Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways. Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical two-state model of hybridization/dehybridization over a smooth landscape with no intermediate metastable states. Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.

@@

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization. Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}. The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} \noteb{these references correspond to CG enhanced sampling studies, I can refer to enhanced sampling on AA here instead (\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations}) but I think the CG studies should be mentioned since not much has been done for AA hybridization} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways. These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacements \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami structure \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.

@@

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers.} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature. The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was set, according to 3SPN.2 documentation, to the coordinates of Arnott et al.\ ~\citep{Arnott1976CRCBiology} The dissociated state was generated by shifting one oligomer from the hybridized state by 1 nm in the x, y, and z directions. The magnitude of this shift did not affect the results so long as all native WC bonds were sufficiently separated. \noteb{No energy minimization (no 3spn2 precedent on this). The lammps default emin techniques may not work for this type of CG forcefield. Given the sampling it seems that  1 $\mu$s of equilibration should be sufficient?} Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the system temperature. \addref{Maxwell-Boltzmann -- \noteb{not sure what to use here, usually not cited?}} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances. We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the $(10 \times 10)$ 100-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair. This featurization naturally accounts for intermolecular configurations via the pairwise distances between bases on different strands. \noteb{previous text had suggested we used the 190-dim features, but we actually used $100-45 = 55$-dim features since the intramolecular features contributed no additional kinetic variance (I've updated accordingly)} One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances. Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2. \citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands. VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization. \citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top 5 modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using 5-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA GK210GL GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively \noteb{not sure the best way to explain this succinctly}. (Figure 1 in the Supporting Information). This motivated the construction of a n-dimensional embedding corresponding to n leading SRV modes retained for each sequence.

\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering. The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.

\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM. The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2. We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns. At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively. The identification of $(i-1)$ implied time scales \noteb{(may be 1-2 additional converging, but these correspond to diffusive processes that are irrelevant or unhelpful in determining the hybridization mechanisms. Still, this might come up when looking at the the SI Fig 2} implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates. We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates. These macrostate MSMs constitute our terminal kinetic models. We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}. This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$. We present in SI Figure 3 the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four kinetic models.

@@

The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}. The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO). Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump. T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs. t-HDVE spectra were acquired with a parallel pulse polarization scheme. Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum. t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.

The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a home-built brass jacket. The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket. The T-jump temperature change ($\Delta$T) was set to $\sim$15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam. The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.

@@

\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}

\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well. It is important to note that we do not include dangling end contributions when calculating NN free energies. Although 3SPN.2 can capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the CG model was not parameterized to fully capture the enthalpic contributions of free bases interaction with terminal base pairs. For each macrostate, dangling ends would be expected to have stabilizing effects \noteb{Can show NN with dangling end contributions (original figure) in SI)} and therefore increase state populations compared to what we observe in simulation.}

\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex. We find that inert tail effects may be particularly impact in predicting out-of-register state stability.}

>>>>

\noindent A robust understanding of the sequence-dependent thermodynamics of DNA hybridization has enabled rapid advances in DNA nanotechnology. A fundamental understanding of the sequence-dependent kinetics and mechanisms of hybridization and dehybridization remains comparatively underdeveloped. In this work, we establish new understanding of the sequence-dependent hybridization/dehybridization kinetics and mechanism within a family of self-complementary pairs of 10-mer DNA oligomers by integrating coarse-grained molecular simulation, machine learning of the slow dynamical pathways, data-driven inference of long-time kinetic models, and experimental temperature-jump infrared spectroscopy. For a repetitive ATATATATAT sequence, we resolve a rugged dynamical landscape comprising multiple metsatable states, numerous competing hybridization/dehybridization pathways, and a spectrum of dynamical relaxations. Introduction of a G:C pair at the terminus (GATATATATC) or center (ATATGCATAT) of the sequence reduces the ruggedness of the dynamics landscape by eliminating a number of metastable states and reducing the number of competing dynamical pathways. Only by introducing a G:C pair midway between the terminus and the center to maximally disrupt the repetitive nature of the sequence (ATGATATCAT) do we recover a canonical two-state model of hybridization/dehybridization over a smooth landscape with no intermediate metastable states. Our results establish new understanding of the dynamical richness of sequence-dependent kinetics and mechanisms of DNA hybridization/dehybridization, present a molecular basis with which to understand experimental temperature jump data, and furnish foundational design rules by which to rationally engineer the kinetics and pathways of DNA association and dissociation in burgeoning DNA nanotechnology applications.

@@

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly stimulate DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971RelaxationOligonucleotides, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, but these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices} A number of computational modeling approaches have also been employed to provide molecular-level resolution of hybridization. Simplified lattice models can recapitulate the essential aspects of the hybridization pathways but lack the realism of continuous space representations \citep{Araque2016LatticeCooperativity, Phys2019}. The long time scales associated with hybridization/dehybridization events place them outside the reach of unbiased all-atom molecular dynamics simulations,\citep{Phys2014} but they can be observed by employing enhanced sampling techniques such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling,\citep{Piana2007AtomisticTransition, Zerze2021ThermodynamicsSimulations, Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} or by using elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques, but approaches to rigorously correct the kinetics, particularly under the conditions of high bias necessary for good sampling, are in their infancy.~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics (CGMD) simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to a speedup factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways. These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}.

@@

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers., Bussi2007AccurateDynamics} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature. The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was defined following 3SPN.2 best practices using the crystal structure coordinates of Arnott et al.~\citep{Arnott1976CRCBiology} The dissociated state was generated from the hybridized state by displacing one strand away from the other by 1 nm in each of the $x$, $y$, and $z$ directions. The magnitude of this displacement did not affect the results so long as all native WC bonds were completely broken. Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the temperature of interest. Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances. We adopt intermolecular pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the 10$\times$10=100-element vector of pairwise distances between all bases within the complementary pair of 10-base strands. One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances. Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2. \citep{Sengupta2019AutomatedSelf-assembly} Finally, we took the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands. VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization. \citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top five modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model. Somewhat surprisingly, we observed that augmenting our set of intermolecular distances with intramolecular distances led to no improvement of the VAMP-2 score, indicating that the kinetically-relevant conformational state of the two strands are adequately represented via the intermolecular distances and leading us to employ only intermolecular distances within our featurization.

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 3D embeddings, respectively.

\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering. The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.

\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM. The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda_i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in \blauw{Fig.~S2}. We observe rapid convergence of the implied time scales $t_i$ with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns. For this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively. The identification of $(i-1)$ implied time scales implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates. We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates. These macrostate MSMs constitute our terminal kinetic models. We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}. This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$ should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$. We present in \blauw{Fig.~S3} the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models perform very well in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four models.

@@

The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}. The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO). Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump. T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between (-10) and 10 fs. t-HDVE spectra were acquired with a parallel pulse polarization scheme. Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum. t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.

The sample was placed between two 1 mm thick CaF2 windows separated by a 50 $\mu$m Teflon spacer enclosed in a home-built brass jacket. The initial temperature of the sample was set using a recirculating chiller connected to the brass sample jacket. The T-jump temperature change ($\Delta$T) was set to 15 $^\circ$C for all measurements and monitored using the change in transmission of the \ce{D2O} bend-libration combination band measured in the LO beam. The temperature change was quantified by comparing the change in transmission of the LO beam with a FTIR temperature series of \ce{D2O}.

@@

%\noteb{Wrote up a different approach to this section should we present the NN data with no dangling ends instead. Have not fully pieced it together, but this is the basis idea:}

%\bood{As illustrated in Fig.~\ref{fig:NN_table_no_de}, we see that our macrostate free energies recapitulate NN model predictions reasonably well. It is important to note that we do not include dangling end contributions when calculating NN free energies. Although 3SPN.2 can capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the CG model was not parameterized to fully capture the enthalpic contributions of free bases interaction with terminal base pairs. For each macrostate, dangling ends would be expected to have stabilizing effects \noteb{Can show NN with dangling end contributions (original figure) in SI)} and therefore increase state populations compared to what we observe in simulation.}

%\bood{A well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end are known to destabilize the duplex. We find that inert tail effects may be particularly impact in predicting out-of-register state stability.}
